Practical Hilbert space approximate Bayesian Gaussian processes for probabilistic programming

### 2020-09-07 Paper/manuscript_v2/manuscript_07_02.tex
add label for the y-axis in figure 17
little corrections to section 2.1

The Mat\'ern class of isotropic covariance functions is given by, 
%
\begin{align*}
k_{\nu}(\bm{\tau})&=\alpha \, \frac{2^{1-\nu}}{\Gamma(\nu)}\left(\frac{\sqrt{2\nu}\bm{\tau}}{\ell}\right)^{\!\nu} \! K_{\nu} \left(\frac{\sqrt{2\nu}\bm{\tau}}{\ell}\right),
\end{align*}
%
where $\nu > 0$ is the order the kernel, $K_{\nu}$ the modified Bessel function of the second kind, and the $\ell > 0$ and $\alpha > 0$ are the length-scale and magnitude (marginal variance), respectively, of the kernel. The particular case where $\nu=\infty$ and $\nu=3/2$ are probably the most commonly used kernels \citep{rasmussen2006gaussian}, 
%
\begin{align*}
k_{\infty}(\bm{\tau})&=\alpha \exp \left( -\frac{1}{2} \frac{\bm{\tau}^2}{\ell^2}\right),  \\
k_{\frac{3}{2}}(\bm{\tau})&=\alpha\left(1+\frac{\sqrt{3}\bm{\tau}}{\ell}\right) \exp\! \left( -\frac{\sqrt{3}\bm{\tau}}{\ell}\right). 
\end{align*}
%
The former is commonly known as the squared exponential or exponentiated quadratic covariance function. Assuming the Euclidean distance between observations, $r=|\bm{x}-\bm{x}'|_{L2}=\sqrt{\sum_{i=1}^{D}(x_i-x_i')^2}$, the kernels written above take the form
%
\begin{align*}
k_{\infty}(r) &= \alpha \exp \left( -\frac{1}{2} \sum_{i=1}^{D}\frac{(x_i-x_i')^2}{\ell_i^2}\right),\\
k_{\frac{3}{2}}(r) &= \alpha \left(1+\sqrt{\sum_{i=1}^{D}\frac{3(x_i-x_i')^2}{\ell_i^2}}\;\right) \exp\! \left( - \sqrt{\sum_{i=1}^{D}\frac{3(x_i-x_i')^2}{\ell_i^2}}\;\right).
\end{align*}
%
Notice that the previous expressions, $k_{\infty}(r)$ and $k_{\frac{3}{2}}(r)$, have been easily generalized to using a multidimensional length-scale $\bm{\ell}\in {\rm I\!R}^D$. Using individual length-scales for each dimension turns the isotropic covariance function into a non-isotropic covariance function. That is, for a non-isotropic covariance function, the smoothness may vary across different input dimensions.

Stationary covariance functions can be represented in terms of their spectral densities \citep[see, e.g.,][]{rasmussen2006gaussian}. In this sense, the covariance function of a stationary process can be represented as the Fourier transform of a positive finite measure \citep[\textit{Bochner's theorem}; see, e.g., ][]{akhiezer1993theory}. If this measure has a density, it is known as the spectral density of the covariance function, and the covariance function and the spectral density are Fourier duals, known as the \textit{Wiener-Khintchine theorem} \citep{rasmussen2006gaussian}. The spectral density functions associated with the Mat\'ern class of covariance functions are given by
%
\begin{align*}
S_{\nu}(\bm{\omega})&= \alpha \, \frac{2^D\pi^{D/2}\Gamma(\nu+D/2)(2\nu)^{\nu}}{\Gamma(\nu)\, \ell^{2\nu}}\left(\frac{2\nu}{\ell^2}+4\pi^2\bm{\omega}^2 \right)^{(\nu+D/2)}
\end{align*}
%
in $D$ dimensions, where vector $\bm{\omega}\in {\rm I\!R^D}$ is in the frequency domain, and $\ell$ and $\alpha$ are the length-scale and magnitude (marginal variance), respectively, of the kernel. The particular cases, where $\nu=\infty$ and $\nu=3/2$, take the form
%
\begin{align}
S_{\infty}(\bm{\omega})&= \alpha\,(\sqrt{2\pi})^D  \ell^D \exp(-\ell^2 \bm{\omega}^2 / 2), \label{eq_specdens_inf}  \\
S_{\frac{3}{2}}(\bm{\omega})&= \alpha\,\frac{2^D\pi^{D/2}\Gamma(\frac{D+3}{2}){3}^{3/2}}{\frac{1}{2}\sqrt{\pi}\ell^3}\left(\frac{3}{\ell^2}+\bm{\omega}^2 \right)^{-\frac{D+3}{2}}. \label{eq_specdens_32} 
\end{align}
%
For instance, with input dimensionality $D=3$ and $\bm{\omega}=(\omega_1,\omega_2,\omega_3)$, the spectral densities written above take the form
%
\begin{align*}
S_{\infty}(\bm{\omega})&= \alpha \, (2\pi)^{3/2}  \prod_{i=1}^{3} \ell_i  \exp\!\left(-\frac{1}{2} \sum_{i=1}^{3} \ell_i^2 \omega_i^2 \right),   \\
S_{\frac{3}{2}}(\bm{\omega})&= \alpha \, 32\,\pi {3}^{3/2}\prod_{i=1}^{3}\ell_i\left(3+\sum_{i=1}^{3}\ell_i^2 \omega_i^2 \right)^{-3},
\end{align*}
%
where individual length-scales $\ell_i$ for each frequency dimension $\omega_i$ have been used.

@@

We do not have a full observation model, as we do not have a model for the censoring process. We use the complementary cumulative log-normal probability distribution for the censored data conditionally on the censoring time $t_i$:
%
\begin{equation*}
p(y_i > t_i \mid f)= \int_{t_i}^{\infty} \LogNormal(y_i \mid f(\bm{x}_i),\sigma^2) \,\mathrm{d}y_i=  1 - \Phi\! \left( \frac{\log(y_i)-f(\bm{x}_i)}{\sigma} \right),
\end{equation*}
%
where $y_i>t_i$ denotes the unobserved survival time.
The latent function $f(\cdot)$ is modeled as a Gaussian process, centered around a linear model of the predictors $\bm{x}$, and with a squared exponential covariance function $k$. Due to the predictor \textit{sex} ($x_2$) being a categorical variable (`1' for female and `2' for male), we apply indicator variable coding for the GP functions, in a similar way such coding is applied in linear models \citep{Gelman+Hill+Vehtari:2020:ROS}. The latent function $f(\bm{x})$, besides of being centered around a linear model, is composed of a general mean GP function, $h(\bm{x})$, defined for all observations, plus a second GP function, $g(\bm{x})$, that only applies to one of the predictor levels ('male' in this case) and is set to zero otherwise:
%
\begin{align*}
h(\bm{x}) &\sim \GP\left(\bm{0}, \, k(\bm{x},\bm{x}', \theta_0) \right),  \\[1pt] 
%
g(\bm{x}) &\sim \GP\left(\bm{0}, \, k(\bm{x},\bm{x}', \theta_1)\right),\\[1pt] 
%
f(\bm{x}) &= c + \bm{\beta} \bm{x} + h(\bm{x}) + \mathbb{I}\!\left[x_2=2\right]g(\bm{x}),
\end{align*}
%
where $\mathbb{I}\left[\cdot\right]$ is an indicator function. Above, $c$ and $\bm{\beta}$ are the intercept and vector of coefficients, respectively, of the linear model. $\theta_0$ contains the hyperparameters $\alpha_0$ and $\ell_0$ which are the marginal variance and length-scale of the general mean GP function, and $\theta_1$ contains the hyperparameters $\alpha_1$ and $\ell_1$ which are the marginal variance and length-scale, respectively, of a GP function specific to the male sex ($x_2=2$). Scalar length-scales, $l_0$ and $l_1$, are used in both multivariate covariance functions, assuming isotropic functions.
%
\begin{figure}
\centering
\includegraphics[scale=0.70, trim = 0mm 5mm 0mm 0mm, clip]{ch5_fig21_posteriors_leukemia.pdf}
\caption{Expected lifetime conditional comparison for each predictor with other predictors fixed to their mean values. The thick line in each graph is the posterior mean estimated using a HSGP model, and the thin lines represent pointwise 95\% credible intervals.}
  \label{ch5_fig21_posteriors_leukemia}
\end{figure}

>>>>

The Mat\'ern class of isotropic covariance functions is given by, 
%
\begin{align*}
k_{\nu}(r)&=\alpha \, \frac{2^{1-\nu}}{\Gamma(\nu)}\left(\frac{\sqrt{2\nu}r}{\ell}\right)^{\!\nu} \! K_{\nu} \left(\frac{\sqrt{2\nu}r}{\ell}\right),
\end{align*}
%
where $\nu > 0$ is the order the kernel, $K_{\nu}$ the modified Bessel function of the second kind, and the $\ell > 0$ and $\alpha > 0$ are the length-scale and magnitude (marginal variance), respectively, of the kernel. The particular case where $\nu=\infty$ and $\nu=3/2$ are probably the most commonly used kernels \citep{rasmussen2006gaussian}, 
%
\begin{align*}
k_{\infty}(r)&=\alpha \exp \left( -\frac{1}{2} \frac{r^2}{\ell^2}\right),  \\
k_{\frac{3}{2}}(r)&=\alpha\left(1+\frac{\sqrt{3}r}{\ell}\right) \exp\! \left( -\frac{\sqrt{3}r}{\ell}\right). 
\end{align*}
%
The former is commonly known as the squared exponential or exponentiated quadratic covariance function. Assuming the Euclidean distance between observations, $r=|\bm{x}-\bm{x}'|_{L2}=\sqrt{\sum_{i=1}^{D}(x_i-x_i')^2}$, the kernels written above take the form
%
\begin{align*}
k_{\infty}(r) &= \alpha \exp \left( -\frac{1}{2} \sum_{i=1}^{D}\frac{(x_i-x_i')^2}{\ell_i^2}\right),\\
k_{\frac{3}{2}}(r) &= \alpha \left(1+\sqrt{\sum_{i=1}^{D}\frac{3(x_i-x_i')^2}{\ell_i^2}}\;\right) \exp\! \left( - \sqrt{\sum_{i=1}^{D}\frac{3(x_i-x_i')^2}{\ell_i^2}}\;\right).
\end{align*}
%
Notice that the previous expressions, $k_{\infty}(r)$ and $k_{\frac{3}{2}}(r)$, have been easily generalized to using a multidimensional length-scale $\bm{\ell}\in {\rm I\!R}^D$. Using individual length-scales for each dimension turns the isotropic covariance function into a non-isotropic covariance function. That is, for a non-isotropic covariance function, the smoothness may vary across different input dimensions.

Stationary covariance functions can be represented in terms of their spectral densities \citep[see, e.g.,][]{rasmussen2006gaussian}. In this sense, the covariance function of a stationary process can be represented as the Fourier transform of a positive finite measure \citep[\textit{Bochner's theorem}; see, e.g., ][]{akhiezer1993theory}. If this measure has a density, it is known as the spectral density of the covariance function, and the covariance function and the spectral density are Fourier duals, known as the \textit{Wiener-Khintchine theorem} \citep{rasmussen2006gaussian}. The spectral density functions associated with the Mat\'ern class of covariance functions are given by
%
\begin{align*}
S_{\nu}(\bm{\omega})&= \alpha \, \frac{2^D\pi^{D/2}\Gamma(\nu+D/2)(2\nu)^{\nu}}{\Gamma(\nu)\, \ell^{2\nu}}\left(\frac{2\nu}{\ell^2}+4\pi^2\bm{\omega}^\intercal \bm{\omega} \right)^{(\nu+D/2)}
\end{align*}
%
in $D$ dimensions, where vector $\bm{\omega}\in {\rm I\!R^D}$ is in the frequency domain, and $\ell$ and $\alpha$ are the length-scale and magnitude (marginal variance), respectively, of the kernel. The particular cases, where $\nu=\infty$ and $\nu=3/2$, take the form
%
\begin{align}
S_{\infty}(\bm{\omega})&= \alpha\,(\sqrt{2\pi})^D  \ell^D \exp(-\frac{1}{2}\ell^2 \bm{\omega}^\intercal \bm{\omega}), \label{eq_specdens_inf}  \\
S_{\frac{3}{2}}(\bm{\omega})&= \alpha\,\frac{2^D\pi^{D/2}\Gamma(\frac{D+3}{2}){3}^{3/2}}{\frac{1}{2}\sqrt{\pi}\ell^3}\left(\frac{3}{\ell^2}+\bm{\omega}^\intercal \bm{\omega} \right)^{-\frac{D+3}{2}}. \label{eq_specdens_32} 
\end{align}
%
For instance, with input dimensionality $D=3$ and $\bm{\omega}=(\omega_1,\omega_2,\omega_3)^\intercal$, the spectral densities written above take the form
%
\begin{align*}
S_{\infty}(\bm{\omega})&= \alpha \, (2\pi)^{3/2}  \prod_{i=1}^{3} \ell_i  \exp\!\left(-\frac{1}{2} \sum_{i=1}^{3} \ell_i^2 \omega_i^2 \right),   \\
S_{\frac{3}{2}}(\bm{\omega})&= \alpha \, 32\,\pi {3}^{3/2}\prod_{i=1}^{3}\ell_i\left(3+\sum_{i=1}^{3}\ell_i^2 \omega_i^2 \right)^{-3},
\end{align*}
%
where individual length-scales $\ell_i$ for each frequency dimension $\omega_i$ have been used.

@@

We do not have a full observation model, as we do not have a model for the censoring process. We use the complementary cumulative log-normal probability distribution for the censored data conditionally on the censoring time $t_i$:
%
\begin{equation*}
p(y_i > t_i \mid f)= \int_{t_i}^{\infty} \LogNormal(y_i \mid f(\bm{x}_i),\sigma^2) \,\mathrm{d}y_i=  1 - \Phi\! \left( \frac{\log(y_i)-f(\bm{x}_i)}{\sigma} \right),
\end{equation*}
%
where $y_i>t_i$ denotes the unobserved survival time.
The latent function $f(\cdot)$ is modeled as a Gaussian process, centered around a linear model of the predictors $\bm{x}$, and with a squared exponential covariance function $k$. Due to the predictor \textit{sex} ($x_2$) being a categorical variable (`1' for female and `2' for male), we apply indicator variable coding for the GP functions, in a similar way such coding is applied in linear models \citep{Gelman+Hill+Vehtari:2020:ROS}. The latent function $f(\bm{x})$, besides of being centered around a linear model, is composed of a general mean GP function, $h(\bm{x})$, defined for all observations, plus a second GP function, $g(\bm{x})$, that only applies to one of the predictor levels ('male' in this case) and is set to zero otherwise:
%
\begin{align*}
h(\bm{x}) &\sim \GP\left(\bm{0}, \, k(\bm{x},\bm{x}', \theta_0) \right),  \\[1pt] 
%
g(\bm{x}) &\sim \GP\left(\bm{0}, \, k(\bm{x},\bm{x}', \theta_1)\right),\\[1pt] 
%
f(\bm{x}) &= c + \bm{\beta} \bm{x} + h(\bm{x}) + \mathbb{I}\!\left[x_2=2\right]g(\bm{x}),
\end{align*}
%
where $\mathbb{I}\left[\cdot\right]$ is an indicator function. Above, $c$ and $\bm{\beta}$ are the intercept and vector of coefficients, respectively, of the linear model. $\theta_0$ contains the hyperparameters $\alpha_0$ and $\ell_0$ which are the marginal variance and length-scale of the general mean GP function, and $\theta_1$ contains the hyperparameters $\alpha_1$ and $\ell_1$ which are the marginal variance and length-scale, respectively, of a GP function specific to the male sex ($x_2=2$). Scalar length-scales, $l_0$ and $l_1$, are used in both multivariate covariance functions, assuming isotropic functions.
%
\begin{figure}
\centering
\includegraphics[scale=0.70, trim = 0mm 5mm 0mm 0mm, clip]{ch5_fig21_posteriors_leukemia_01.pdf}
\caption{Expected lifetime conditional comparison for each predictor with other predictors fixed to their mean values. The thick line in each graph is the posterior mean estimated using a HSGP model, and the thin lines represent pointwise 95\% credible intervals.}
  \label{ch5_fig21_posteriors_leukemia}
\end{figure}
